Speed and Precision in Range Analysis

### 2012-05-16 doc/Manuscript/SPE/paper.tex
Added Vic's new chart.

\documentclass[11pt]{article}

% Packages that I must import to compile this latex file:
\usepackage{amssymb}    % To use \mathbb
\usepackage{amsmath}    % To use the equation environment
\usepackage{graphicx}   % For includegraphics
\usepackage{mathpartir} % To use \inferrule

@@

\title{Global Range Analysis}
\author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
Fernando Magno Quint\~{a}o Pereira}
\date{}
\maketitle

@@

@@

% The context: why range analysis is necessary.
The analysis of integer variables on the interval lattice has been the
canonical example of abstract interpretation since its introduction in
Cousot and Cousot's seminal paper~\cite{Cousot77}.
Compilers use range analysis to infer the possible values that discrete
variables may assume during program execution.
This analysis has many uses.
For instance, it allows the optimizing compiler to remove from the program text
redundant overflow tests~\cite{Sol11} and unnecessary array bound
checks~\cite{Bodik00}.
Additionally, range analysis is essential not only to the bitwidth aware
register allocator~\cite{Barik06,Tallam03}, but also to more traditional
allocators that handle registers of different
sizes~\cite{Kong98,Pereira08,Scholz02}.
Additionally, range analysis has also been used to statically predict the
outcome of branches~\cite{Patterson95}, to detect buffer overflow
vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
loops~\cite{Lokuciejewski09}
and even to synthesize hardware~\cite{Cong05,Lhairech10,Mahlke01}.

@@

\begin{figure}[hbt]
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{0.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
A suite of constraints that produce an instance of the range analysis problem.}
\end{small}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
\end{center}
\caption{\label{fig:timeComp}
Runtime comparison between intra, inter and inter+inline versions of
our algorithm.
The bars are normalized to the time to run the intra-procedural analysis.
}
\end{figure}

@@

@@

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit they use domain specific tools.
Astr\'{e}, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
they do not provide results about its precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.

@@

\bibliographystyle{plain}
\bibliography{../references}

>>>>

\documentclass[times]{speauth}

% Packages that I must import to compile this latex file:
\usepackage{amssymb}    % To use \mathbb
\usepackage{amsmath}    % To use the equation environment
\usepackage{graphicx}   % For includegraphics
\usepackage{mathpartir} % To use \inferrule
\usepackage{moreverb}   % Requirement from SPE. See if you can remove it.

\usepackage[dvips,colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlco
lor=red]{hyperref}      % Anther requirement from SPE.

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2012}

@@

\runningheads{V. H. S. Campos {\em et al.}}{Speed and Precision in Range Analysis}

\title{Speed and Precision in Range Analysis}

\author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
Fernando Magno Quint\~{a}o Pereira\corrauth}

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil}

\corraddr{fernando@dcc.ufmg.br}

@@

\maketitle

@@

% The context: why range analysis is necessary.
The analysis of integer variables on the interval lattice has been the
canonical example of abstract interpretation since its introduction in
Cousot and Cousot's seminal paper~\cite{Cousot77}.
Compilers use range analysis to infer the possible values that discrete
variables may assume during program execution.
This analysis has many uses.
For instance, it allows the optimizing compiler to remove from the program text
redundant overflow tests~\cite{Sol11} and unnecessary array bound
checks~\cite{Bodik00,Gampe11}.
Additionally, range analysis is essential not only to the bitwidth aware
register allocator~\cite{Barik06,Tallam03}, but also to more traditional
allocators that handle registers of different
sizes~\cite{Kong98,Pereira08,Scholz02}.
Additionally, range analysis has also been used to statically predict the
outcome of branches~\cite{Patterson95}, to detect buffer overflow
vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
loops~\cite{Lokuciejewski09}
and even to synthesize hardware~\cite{Cong05,Lhairech10,Mahlke01}.

@@

\begin{figure}[hbt]
\begin{center}
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{1.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
A suite of constraints that produce an instance of the range analysis problem.}
\end{small}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
\includegraphics[width=0.485\textwidth]{images/timeWideningOperators}
\end{center}
\caption{\label{fig:timeComp}
(Left) Runtime comparison between intra, inter and inter+inline versions of
our algorithm.
(Right) Runtime comparison between different widening operators.
The bars are normalized to the time to run the intra-procedural analysis.
}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By combining the different widening operators -- simple or jump-set -- with
the different number of pre-evaluations -- in our case 0 or 16, we have four
different widening strategies.
Figure~\ref{fig:timeComp}(Right) compares the runtime of all these strategies
for the integer programs in the SPEC CPU 2006 collection.
We have observed no measurable different between the jump-set and the simple
operator: the latter is 0.93\% faster, e.g., 1.0093x faster.
The strategy that precedes the simple operator with 16 rounds of pre-evaluation
is 7\% slower than the strategy that does not do any pre-evaluation.
Finally, the combination of 16 rounds of pre-evaluation, plus jump-set widening
is 13\% slower than the simple widening strategy.
We observe an anomalous behavior in \texttt{astar} and \texttt{mcf}: the simple
strategy results in a small slowdown.
These benchmarks have the two fastest runtimes in the benchmark suite; e.g.,
we analyze \texttt{mcf} in 0.02 seconds.
Thus, we believe that the unexpected behavior is due to runtime noise that
outlives a sequence of 100 executions of each benchmark.

@@

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit they use domain specific tools.
Astr\'{e}, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
they do not provide results about its precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.

@@

\bibliographystyle{wileyj}
\bibliography{../references}
